\begin{acknowledgements*}
I would like to thank my family for their great support, especially my brother, who always encouraged me, even when this thesis did not progress in the way it should.

A big thank you to my supervisor Markus Raab, who was always there, when I needed a helping hand. His kind words and support were an integral part of the process of writing and finishing this thesis. I would also like to thank Franz Puntigam for giving me the chance to work on this interesting research topic.

Thank you to all the people who helped me in the preparation and implementation phase of the thesis. I greatly appreciate the work of the other Elektra developers, be it documentation text or support code. Your work was integral in helping me to implement the parsing code. A special thank to all the Elektra developers that took part in the discussion about YAML features, and helped me decide which YAML subset the parsing code should support. I also like to thank the communities of \glstext{PEGTL} and \glstext{YAEP}, which implemented features and bug fixes that eased the implementation phase of the thesis considerably.

Last but not least, I would like to thank my friend Johannes Eder for his input in the implementation phase, debugging assistance, and for keeping me company while I worked on this thesis.
\end{acknowledgements*}

\vfill
\begin{leftbar}
\textbf{Note:} We use hyperlinks quite heavily in this thesis. This makes it easier for you to quickly jump between relevant parts of the text and access various online resources. Please consider this advantage before you print this text.
\end{leftbar}
